---
title: "Haplogroups classification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Haplogroups classification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this tutorial, I will show how to use *mitor* and *haplogrep* to classify haplogroups of downloaded sequences. In particular, I will show:

- How to fetch some mitochondrial sequences from the NCBI database.
- How to classify their haplogroups using *haplogrep*.
- How to sample a given number *n* sequences for each haplogroup.

## Fetching mtDNA sequences

Here, I will query the [NCBI Nucleotide database](https://www.ncbi.nlm.nih.gov/nucleotide/) to retrieve all the *Homo Sapiens* mtDNA sequences.

First, we load the *mitor* package.

```{r loading, eval=FALSE}
library(mitor)
```

After the package is loaded, we now define a query to fetch *Homo Sapiens* mtDNA sequences from the [NCBI Nucleotide database](https://www.ncbi.nlm.nih.gov/nucleotide/). For more information regarding the query structure check their [website](https://www.ncbi.nlm.nih.gov/books/NBK44863/).

```{r query, eval=FALSE}
query <-
  "(016500[SLEN]:016600[SLEN]) AND mitochondrion[FILT] AND txid9606[orgn] NOT txid63221[orgn] NOT txid741158[orgn] NOT NC_012920.1[accn]" 
```

The query I am using has several filters:

- **(016500[SLEN]:016600[SLEN])** specifies that I want only sequences with length between 16500 and 16600 bases.
- **mitochondrion[FILT]** specifies to include only mitochondrial sequences.
- **txid9606[orgn]** specifies that the sequences have to belong to *Homo Sapiens*.
- **NOT txid63221[orgn] NOT txid741158[orgn]** specifies to not include the *Homo Sapiens* subspecies *Neanderthals* and *Denisovans*.
- **NOT NC_012920.1[accn]** specifies to not include the revised Cambridge Reference Sequence, since it already comes with *mitor*.

Now, we can retrieve the sequences using the *fetch_seq* function. To fetch all the sequences corresponding to the query it is enough to not specify any *n*.

```{r fetch, eval=FALSE}
# Define name of direcotry where to save the sequences
dir_path <- "data"
# Define the filename
filename <- "hs_mtDNA"
fetch_seq(query = query,
          dir_path = dir_path,
          filename = filename)
```
