---
title: "Haplogroups classification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Haplogroups classification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this tutorial, I will show how to use *mitor* and *haplogrep* to classify haplogroups of downloaded sequences. In particular, I will show:

- How to fetch some mitochondrial sequences from the NCBI database.
- How to classify their haplogroups using *haplogrep*.
- How to sample a given number *n* sequences for each haplogroup.

## Fetching mtDNA sequences

Here, I will query the [NCBI Nucleotide database](https://www.ncbi.nlm.nih.gov/nucleotide/) to retrieve all the *Homo Sapiens* mtDNA sequences.

First, we load the *mitor* package.

```{r loading, eval=FALSE}
library(mitor)
```

After the package is loaded, we now define a query to fetch *Homo Sapiens* mtDNA sequences from the [NCBI Nucleotide database](https://www.ncbi.nlm.nih.gov/nucleotide/). For more information regarding the query structure check their [website](https://www.ncbi.nlm.nih.gov/books/NBK44863/).

```{r query, eval=FALSE}
query <-
  "(016500[SLEN]:016600[SLEN]) AND mitochondrion[FILT] AND txid9606[orgn] NOT txid63221[orgn] NOT txid741158[orgn] NOT NC_012920.1[accn]" 
```

The query I am using has several filters:

- **(016500[SLEN]:016600[SLEN])** specifies that I want only sequences with length between 16500 and 16600 bases.
- **mitochondrion[FILT]** specifies to include only mitochondrial sequences.
- **txid9606[orgn]** specifies that the sequences have to belong to *Homo Sapiens*.
- **NOT txid63221[orgn] NOT txid741158[orgn]** specifies to not include the *Homo Sapiens* subspecies *Neanderthals* and *Denisovans*.
- **NOT NC_012920.1[accn]** specifies to not include the revised Cambridge Reference Sequence, since it already comes with *mitor*.

Now, we can retrieve the sequences using the *fetch_seq* function. To fetch all the sequences corresponding to the query it is enough to not specify any *n*.

```{r fetch, eval=FALSE}
# Define name of direcotry where to save the sequences
dir_path <- "data"
# Define the filename
filename <- "hs_mtDNA"
fetch_seq(query = query,
          dir_path = dir_path,
          filename = filename)
```

## Running haplogrep

To classify the haplogroups of the downloaded sequences, we first need to load the sequences.

```{r haplogrep, eval=FALSE}
my_seqs <- Biostrings::readDNAStringSet(paste0(dir_path, filename))
```

At this point, we can simply run the function *classify_haplogroup* on the loaded sequences. Sequences containing IUPAC ambiguities characters not accepted by *haplogrep* will be removed.

```{r, eval=FALSE}
haplogroups_df <- classify_haplogroup(my_seqs)
```

The resulting data frame looks like this:

        SampleID   Haplogroup Rank Quality             Range
    1   DQ856316.1        HV1b2    1  1.0000           1-16569
    2   EU603401.1        K1a4a    1  0.9272           1-16569
    3   KX681447.1          H7e    1  0.9610           1-16569
    4   KX702230.1           H7    1  0.8705           1-16569

*SampleID* corresponds to the name of the sequence, *Haplogroup* to the haplogroup assigned to it by *haplogrep*, *Quality* to the confidence of the classification. Do not consider *Rank* and *Range*.
We can now use this table to sample sequences according to the haplogroup.

## Sampling sequences

If we are, for example, interesting in sampling 5 sequences per haplogroup we can proceed as follow.

```{r, eval=FALSE}
# Filter sequences according to the quality of classification (arbitrary)
haplogroups_df <- haplogroups_df %>% dplyr::filter(Quality>0.85)
# Consider only the macro haplogroup (first letter)
haplogroups_df$Haplogroup <- as.factor(strtrim(haplogroups_df$Haplogroup, 1))
# Sample 5 sequences per haplogroup (if available)
sampled_seqs_df <- haplogroups_df %>%
  dplyr::group_by(Haplogroup) %>%
  dplyr::slice_sample(n = min(n(), 5))
# Subset the my_seqs
sampled_seqs <- my_seqs[sampled_seqs_df$SampleID]

```

Now, *sampled_seqs* contains 5 sequences (if available) per haplogroup. It  is up to you how you can use them!

